Indigenous Subsistence and Colonial Encounters: A Zooarchaeological Perspective from Kiwulan in Northeastern Taiwan

### 2025-03-06 analysis/paper/paper.qmd
edit the introduction section and add references

---
title: "Indigenous Subsistence and Colonial Encounters: A Zooarchaeological Perspective from Kiwulan, a late Iron site in Northeastern Taiwan"
author:
  - Li-Ying Wang:
      correspondence: "yes"
      email: liyingw15@gmail.com
      orcid: 0000-0001-9060-2125
      institute: Academia Sinica
title-block-published: "Last updated"
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Colonial encounters between Europeans and Indigenous people often led to profound impacts on many Indigenous societies and cultures around the world. In northeastern Taiwan, local Indigenous groups experienced foreign presences through regional trade networks in East Asia, which later expanded into a larger-scale trade network after the arrival of Europeans in the 17th century. This research focuses on local subsistence through faunal remains to understand how Indigenous communities responded to the regional trade activities driven by the European and Han Chinese, particularly the high demand for natural resources, such as deer. By analyzing faunal remains from the upper cultural layer of Kiwulan (AD 1450-1950), a late Iron Age site in northeastern Taiwan, the local participation in deer hide trade and their adaptive strategies were discussed. The results suggest indigenous cultural persistence of deer consumption practiced by local people, along with increased reliance on suids during the European period as the demands for deer products intensified. This paper also provides insights into colonial experiences from Indigenous settlements at the periphery of European territories, a perspective rarely discussed.

keywords: |
  Zooarchaeology; Iron Age; Taiwan; sika deer; trade; European colonization
highlights: |
  These are the highlights. 
---

@@

# Introduction
<!-- introduction about European expansion -->
European expansion since the 15th century was usually associated with long-distance trade, exploration, and colonization in many parts of the world, which often led to profound impacts on indigenous societies. Depending on particular geographic and historical circumstances, many regions experienced different forms of European colonialism and power relations [@Jordan2009]. Understanding the dynamics of these colonial processes has become a significant issue in archaeology because it provides critical insights into the complex outcomes of European colonization [@Gosden2004]. In Asia, European colonization was mainly motivated by the acquisition of raw materials, natural resources, or luxury items from Asia for commercial purposes [@Prakash1997]. Although the new exchange systems or colonial institutions introduced by Europeans could change local economic systems, social structures, and settlement patterns [@Christie1996; @Junker2004; @Lape2002], indigenous groups did not just accept European colonization passively. Recent discussions on colonial interactions also suggest the active role of indigenous groups in responding to the arrival of Europeans and presenting resistance, negotiation, persistence, resilience, and entanglement [@Torrence2003; @Silliman2005; @Mitchell2010; @Jordan2009; @Alexander1998; @Noe2023].

<!-- NE Taiwan as a case study -->
While many studies focus on prominent colonies, places on the periphery of colonial centers were also more and less impacted by European powers or associated trading systems [@Acabado2017; @Trabert2017]. Exploring these areas is important to gain a broader understanding of the varied consequences of colonial expansion. Yilan, located in northeastern Taiwan, is an example of a periphery region where European control was limited and indirect for most of the time but intensively involved in regional trade activities stimulated by the European presence. In the mid-17th century, the VOC (Dutch East India Company) formalized the deerskin trade in Taiwan for export on a large scale, involving many Indigenous communities and Han Chinese settlers [@Heyns2001; @Cheng2017]. Despite some discussions on the operations and policies of the deerskin trade, direct investigations on archaeological evidence into the experiences of local indigenous people with the commercial activities are rare. The European impacts on local ecological conditions have been recorded in many places in colonial settings [@Jordan2009]. Whether the interaction between Indigenous people and their environment changed due to European trade demands requires further archaeological investigation.

<!-- research aim and questions -->
In this paper, I will explore the interactions between people and their environment in a colonial context by examining the resource use of indigenous people through a case study at the Kiwulan site (1350-1950 CE), the largest Iron Age settlement in northeastern Taiwan [@Chen2007]. Kiwulan encountered European colonial power in the 17th century, followed by a large influx of Han Chinese migration in the 19th century. What was the resource use of Indigenous people before and after the foreign presence? Can we identify subsistence strategies in response to impacts resulting from integration into trading network? To address these questions, I will analyze the zooarchaeological assemblages from Kiwulan, because they provide valuable insights into subsistence practices and accommodation within specific social contexts, such as European colonization or being part of the regional trade network [@Defrance2009; @Crabtree1990]. Through a perspective from subsistence, this study aims to shed light on Indigenous persistence and resilience in the face of foreign presence.

# European Encounters in the Northern Taiwan
<!-- Spanish colonization and the power relations in particular spatial and temporal contexts-->
In 1626, the Spanish East Indies Company (Indias Orientales Españolas) arrived in northern Taiwan and established the San Salvador colony at Heping Dao in Keelung, followed by Santo Domingo in Tamsui, to counter the threat of the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) in the Asian market, which had established a colony in southern Taiwan two years earlier [@Borao2009, p.41; @Andrade2007]. Although the Spanish intended to build a colonial city, San Salvador remains its early development during their colonization, primarily consisting of a fortress and a church [@Borao2009, pp. 103-134; @Berrocal2018]. The neighboring area of the fort was inhavited by local indigenous groups, the Taparris and the Quimaurris, who had lived at the location of the fort before being expelled by the Spanish [@Borao2001, SIT:165, p.81]. The relations between the Spanish and indigenous people might not involve extensive colonialism or intercultural domination, and the Spanish territorial control could be limited to the fort. Instead, the Spanish relied on the exchange of local goods and food surpluses with local indigenous people, according to written records [@Borao2001, SIT:170-171]. This corresponds with a higher proportion of indigenous and Chinese items with little European material found in San Salvador [@Berrocal2018], indicating cultural entanglement.

@@

# Archaeological Context at Kiwulan
<!--introduction of Kiwulan and the population-->
Kiwulan, located in the northern part of Yilan, is a multi-component site, where the upper component spans from the late Iron Age to the historical periods (AD 1350-1850) [@Chen2007]. As a salvage project, a large scale excavation was conducted at Kiwulan to recover information before it was destroyed. A total of 262 excavation units uncovered abundant artifacts and features, including middens, burials, and post-holes with wooden posts in situ. It suggests that Kiwulan is a well-developed settlement with longhouses (approximately 40 x 80 m) situated side by side and burials located to the north of the residential area [@Chen2004]. The main inhabitants of Kiwulan were the Kavalan people, as indicated by local oral history and the similarities between Kavalan material culture recorded in historical records and archaeological artifacts [@Chen2004]. According to a mid-17th-century Dutch census, there were about 40 indigenous villages with approximately 10,000 people, and Kiwulan (referred to as Kibannoran in Dutch records) was the largest village, with 160 households and 711 people [@Nakamura1938]. In the early 19th century, the population of Kiwulan dramatically decreased to fewer than 100 people due to the population pressure of Han Chinese migrants, which led to the relocation of Kavalan people to eastern Taiwan [@Zhan1995].

@@

@@

Since deer is the most frequent animal at Kiwulan, the frequency of skeletal parts was examined to explore any changes in processing decisions of meat portions between different phases. Deer bones were grouped into six body portions based on their nutritional values [@Madrigal2004], including head (cranium, mandible), axial (vertebrae, sternum, ribs), upper forelimb (scapula, humerus, ulna, and radius), upper hindlimb (femur, tibia, innominate), lower limb (carpals, metacarpals, tarsals, metatarsals, astragalus, calcaneus), and foot (phalanges). The result (@fig-deer-portion: A) shows that the meaty portion, especially the upper forelimb, is the most frequent part from CL6 to CL2, followed by upper hindlimb and the less-meaty portions, lower limb and foot. The consistency of higher numbers of upper limb and lower limb suggests that those elements should be transported together (e.g., @Stiner1991). The genereally low frequency of axial element, such as vertebrae and ribs, might be likely due to a high degree of fragmentation, butchery practices, or dietary preferences. A different pattern between layers is observed in the frequency of cranial bones, for which a few present in CL6, increased to the highest number in CL4, and then decreased thereafter. In addition, lower limb and foot bones appear to be proportionally less in CL2 and CL3 compared to earlier periods.

@@

# Discussion
<!--results and chronology -->
The sampled faunal assemblage from the major residential area at Kiwulan suggests a continuation of sika deer consumption from CL6 to CL2, and then shifted to cattle in CL1. Despite deer was a substantial part of dietary practices at Kiwulan, variations in their quantity, ubiquity, butchering patterns, and proportions relative to other animals show changes that could reflect the indigenous reactions to foreign presence. Based on radiocarbon dates and diagnostic artifacts indicating specific time [@Wang2020ornament; @Chen2007], the cultural layers were further grouped into temporal sequences in relation to foreign presences for further discussion. The temporal sequences include the pre-European period (CL6, CL5), the European period (CL4), the post-European period (CL3), and the arrival of Han-Chinese immigrants (CL2, CL1). It should be noted that the European period does not necessarily mean a solid European colonial control over Kiwulan, but a broad impact from European occupation in Northern Taiwan and associated colonial economic activities on indigenous people in Yilan, such as getting involved in the long-distance trade network, tribute paying, and suffered from occasional European attacks.

@@

A consistent pattern was also identified throughout all periods: High frequencies of upper limb bones but low frequencies of vertebrae and ribs, suggesting selectively butchered of meaty portions. Moreover, it is interesting that front quarters (humerus and radius) are much more abundant than hindquarters (femur and tibia). One possible explanation is that femurs and tibiae were heavily fragmented for marrow extraction due to their high yields [@Madrigal2002; @Lupo1998], making them less identifiable in the assemblage. Marrow extraction can be supported by at least 25% of long bones showing fractures and a moderate degree of fragmentation. Another likelihood is that some hindquarters might be traded, as European records mention deer legs and meats as commodities in addition to hides [@Blusse2006, pp.220-224], but without specifying which legs. Considering the overall high proportion of meaty portions at Kiwulan, deer were likely processed not only for local consumption but also for trade.

@@

<!-- population dynamics and arguments-->
The population dynamics at Kiwulan must be carefully considered when interpreting the faunal results, because high population during the European period inferred by pottery could contribute to the increasing animal use overall [@Wang2020pot; @Hsieh2009]. However, the impacts from foreign presence can be still identified through butchery patterns of deer parts, the relative proportions of animals, and taxon ubiquity. First, deer bones with skinning evidence were mostly found during the European period, followed by the late pre-European phase, corresponding with the rise of regional trade networks since the 15th century and intensified trade in the 17th century. Second, if the high frequency of deer parts during the European period, especially heads relative to brains or tongues consumption, solely resulted from food stress due to population growth, a parallel rise in marrow extraction would be expected. In fact, long bones with fracture patterns were 10-30 % lower during the European period  and the previous phase than in other periods. Third, population growth should have led to a more proportional increase in both large and small animals. Although there was a high taxonomic diversity during the European period, suids present a disproportionate rise in their ubiquity, quantity, and proportion. This indicates that trade of deer products may have driven indigenous people to rely more on suids as an alternative prey, likely due to their comparable body size and meat return rates to deer [@Broughton2011prey]. Suids as a secondary protein source after deer are also observed in prehistoric western Taiwan [@Chuang2021]. This hints the role of suids as a supplement when deer was not sufficient.

<!--indigenous reaction and resilience-->
Compelling the zooarchaeological evidence and descriptions from historical records, the substantial increase in deer bones in the European period not just simply the result of population growth, but also reflecting intensified hunting driven by deer trade. The results suggest deer is the first priority of food source and commodities at Kiwulan since the late pre-European period and their economic benefits were intensified during the European period. It should be noted that the role of the Dutch is both colonist who demanded tribute from Yilan, and another trader who were incorporated into the pre-existing inter-insular trade network involving Han Chinese merchants and other indigenous groups in Northern Taiwan [e.g., @Torrence2000]. The prestige values embedded in foreign goods in indigenous culture might lead Kiwulan aggrandizers actively participated in the trade rather than resisted [@Wang2020ornament]. After the Dutch left, deer hide trade continued but back to a small scale, with around 10,000 hides exported annually, which is 10 times less than during the European period [@Nakamura1997]. A small deer trade market and low population at Kiwulan might lead to a reduced reliance of suids, as shown by a low proportion of suids in the post-European period and eventually disappeared during the Chinese period.

<!--Final remarks -->
Everyday practices play an essential role for maintaining Indigenous traditions within a colonial setting or frequent cultural exchange [@Silliman2009; e.g., @Noe2023]. The persistence of deer consumption with similar butchery patterns at Kiwulan reflect a maintained tradition, showing community identities.

while the increased utilization of suids likely responded to the high demand for deer products in large scale trade.  The small sample size after separating them into different periods might limits a further interpretation; however, changes in different phases were still provide evident patterns reflecting changes in animal use. This highlights how Kiwulan residents reacted to foreign presence and demonstrated cultural resilience.

Deer continued to be consumed until the early Chinese period.

# Conclusion
By contextualizing zooarchaeological evidence into time periods relative to foreign presence, including the Europeans in the 17th century, and Chinese immigrants from the late 19th century, this study reveals the participation of indigenous groups in regional trade work. The frequencies, butchering patterns, and human modifications on deer bones confirms the active deer hides trade since the early 15th century and intensified during the European period. The results demonstrates even indigenous people in peripheral areas with relatively weak colonial control were also profoundly impacted than previously thought. In pericolonial setting, the continued prioritization of deer consumption at Kiwulan suggests a cultural persistence of food practices, while the strategic reliance on suids during the European period demonstrates adaptation to high demands for deer products, and later incorporating more muntjac in response to overhunting of sika deer. The long-term dynamics of animal use in northeastern Taiwan can enhance our understanding of Indigenous resilience in face of foreign impacts and environmental change. In addition, the role of indigenous people might be more active in the exchange network considering the large amount of foreign goods that were incorporated into local cultures and used as prestige items. A more comprehensive investigation across Indigenous settlements in northeastern Taiwan and including other dietary evidence, such as botanical remains and organic residue analysis in the future could provide a deeper insights into the regional dietary patterns and human-environment dynamics.

@@

<!-- The following line inserts a page break  -->

>>>>

---
title: "Indigenous Subsistence and Colonial Encounters: A Zooarchaeological Perspective from Kiwulan in Northeastern Taiwan"
author:
  - Li-Ying Wang:
      correspondence: "yes"
      email: liyingw15@gmail.com
      orcid: 0000-0001-9060-2125
      institute: Academia Sinica
title-block-published: "Last updated"
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Colonial encounters between Europeans and Indigenous people often led to profound impacts on many Indigenous societies and cultures around the world. In northeastern Taiwan during the Late Iron Age, local Indigenous groups experienced intense foriegn contacts through due to regional trade networks in East Asia, which later expanded into a larger-scale trade after the arrival of Europeans in the 17th century. This research focuses on local subsistence through faunal remains to understand how Indigenous communities responded to the regional trade activities driven by the Europeans and the Han Chinese, particularly the high demand for natural resources, such as deer. By analyzing faunal remains from the upper cultural layer of Kiwulan (AD 1450-1950), a late Iron Age site in northeastern Taiwan, the local participation in deer hide trade and their adaptive strategies were discussed. The results suggest indigenous cultural persistence of deer consumption, along with increased reliance on suids during the European period as the demands for deer products intensified. This paper also provides insights into colonial experiences from Indigenous settlements at the periphery of European colonial territories, a perspective rarely discussed.

keywords: |
  Zooarchaeology; Iron Age; Taiwan; sika deer; trade; European colonization
highlights: |
  Indigenous community in northeastern Taiwan engaged in regional deer trade were identified from faunal remains.
  The high freqiencies of deer bones at Kiwulan reveals continously cultural persistence in deer consumption.
  Changes in deer butchery patterns and cooking practices indicates an intensified hunting during the European colonization period.
  Increased reliance on suids during the European period suggests adaptive strategies to sustain local subsistence.
---

@@

# Introduction
<!-- introduction about European expansion -->
European expansion since the 15th century was usually associated with long-distance trade, exploration, and colonization in many parts of the world, which often led to profound impacts on indigenous societies. Depending on particular geographic and historical circumstances, many regions experienced different forms of European colonialism and power relations [@Jordan2009]. Understanding the dynamics of these colonial processes has become a significant issue in archaeology because it provides critical insights into the complex outcomes of European colonization [@Gosden2004]. In East Asia, European colonization was mainly motivated by the acquisition of raw materials, natural resources, or luxury items for commercial purposes [@Prakash1997]. The new exchange systems and colonial institutions introduced by Europeans often affected local economic systems, social structures, and settlement patterns [@Christie1996; @Junker2004; @Lape2002]. However, many recent studies also suggest that indigenous groups did not just accept European colonization passively but actively responded to the arrival of Europeans and colonialism through resistance, negotiation, and cultural persistence and resilience, showing entangled colonial experiences [@Torrence2003; @Silliman2005; @Mitchell2010; @Jordan2009; @Alexander1998; @Noe2023].

<!-- NE Taiwan as a case study -->
While many studies focus on prominent colonies, places on the periphery of colonial bases were also more and less impacted by European powers or associated trading systems [@Acabado2017; @Trabert2017]. Exploring these areas is important to gain a broader understanding of the varied consequences of colonial expansion. Yilan, located in northeastern Taiwan, is an example of a remote region where European control was limited and indirect for most of the time but was intensively involved in regional trade activities with foreign groups. In the mid-17th century, the VOC (Dutch East India Company) monopolized the deerskin trade in Taiwan for export on a large scale, involving many Indigenous communities and Han Chinese settlers [@Heyns2001; @Cheng2017]. Despite many discussions on the operations and policies of the deerskin trade through Dutch and Chinese written records [@Lin2014deer], direct investigations on the colonial experiences of local indigenous people through archaeological evidence are rare. The interaction between Indigenous people and their environments in the colonial setting of economic demands requires more Indigenous perspectives from archaeological studies.

<!-- research aim and questions -->
In this paper, I will investigate the Indigenous subsistence in a colonial context through the case study of the Kiwulan site (AD 1350-1950), one of the largest Iron Age settlements in northeastern Taiwan [@Chen2007]. Kiwulan encountered European colonial powers, including the Spanish and Dutch in the 17th century, followed by a large influx of Han Chinese migration in the 19th century, making it a suitable example to explore the reactions of Indigenous groups to the foreign presence over time. My research questions are: What was the resource use of Indigenous people before and after those foreign presences, especially during the Dutch period with large-scale trade? Can we identify the evidence of getting involved in the deer trade from faunal remains? If so, were there subsistence strategies in response to the impacts of economic demands by the Europeans and Chinese? To address these questions, I will analyze the faunal remains with a focus on vertebrate species from Kiwulan because zooarchaeology is one useful approach to studying local consumption and exchange systems [@Defrance2009; @Crabtree1990]. Through zooarchaeological evidence, this study aims to stress the active role of Indigenous people at the periphery of colonial bases in regional trade networks.

# European Encounters in the Northern Taiwan
<!-- Spanish colonization and the power relations in particular spatial and temporal contexts-->
In 1626, the Spanish East Indies Company (Indias Orientales Españolas) arrived in northern Taiwan and established the San Salvador colony at Heping Dao in Keelung, followed by Santo Domingo in Tamsui (@fig-KWL-map), to counter the threat of the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) in the Asian market, which had established a colony in southern Taiwan two years earlier [@Borao2009, p.41; @Andrade2007]. Although the Spanish intended to build a colonial city, San Salvador remains its early development during their colonization, primarily consisting of a fortress and a church [@Borao2009, pp. 103-134; @Berrocal2018]. The neighboring area of the fort was inhavited by local indigenous groups, the Taparris and the Quimaurris, who had lived at the location of the fort before being expelled by the Spanish [@Borao2001, SIT:165, p.81]. The relations between the Spanish and indigenous people might not involve extensive colonialism or intercultural domination, and the Spanish territorial control could be limited to the fort. Instead, the Spanish relied on the exchange of local goods and food surpluses with local indigenous people, according to written records [@Borao2001, SIT:170-171]. This corresponds with a higher proportion of indigenous and Chinese items with little European material found in San Salvador [@Berrocal2018], indicating cultural entanglement.

@@

# Archaeological Context at Kiwulan
<!--introduction of Kiwulan and the population-->
Kiwulan, located in the northern part of Yilan (@fig-KWL-map), is a multi-component site, where the upper component spans from the late Iron Age to the historical periods (AD 1350-1850) [@Chen2007]. As a salvage project, a large scale excavation was conducted at Kiwulan to recover information before it was destroyed. A total of 262 excavation units uncovered abundant artifacts and features, including middens, burials, and post-holes with wooden posts in situ. It suggests that Kiwulan is a well-developed settlement with longhouses (approximately 40 x 80 m) situated side by side and burials located to the north of the residential area [@Chen2004]. The main inhabitants of Kiwulan were the Kavalan people, as indicated by local oral history and the similarities between Kavalan material culture recorded in historical records and archaeological artifacts [@Chen2004]. According to a mid-17th-century Dutch census, there were about 40 indigenous villages with approximately 10,000 people, and Kiwulan (referred to as Kibannoran in Dutch records) was the largest village, with 160 households and 711 people [@Nakamura1938]. In the early 19th century, the population of Kiwulan dramatically decreased to fewer than 100 people due to the population pressure of Han Chinese migrants, which led to the relocation of Kavalan people to eastern Taiwan [@Zhan1995].

@@

```{r}
#| label: fig-KWL-map
#| echo: false
#| fig-cap: "The geographic location of Kiwulan and European fortress in northern Taiwan (adapted from Wang and Marwick, 2020). Map data from naturalearthdata.com."

knitr::include_graphics(here::here("analysis/figures/KWL_map.png"))
```

@@

Since deer is the most frequent animal at Kiwulan, the frequency of skeletal parts was examined to explore any changes in processing decisions of meat portions between different phases. Deer bones were grouped into six body portions based on their nutritional values [@Madrigal2004], including head (cranium, mandible), axial (vertebrae, sternum, ribs), upper forelimb (scapula, humerus, ulna, and radius), upper hindlimb (femur, tibia, innominate), lower limb (carpals, metacarpals, tarsals, metatarsals, astragalus, calcaneus), and foot (phalanges). The result (@fig-deer-portion: A) shows that the meaty portion, especially the upper forelimb, is the most frequent part from CL6 to CL2, followed by upper hindlimb and the less-meaty portions, lower limb and foot. The consistency of higher numbers of upper limb and lower limb suggests that those elements should be transported together [e.g., @Stiner1991]. The genereally low frequency of axial element, such as vertebrae and ribs, might be likely due to a high degree of fragmentation, butchery practices, or dietary preferences. A different pattern between layers is observed in the frequency of cranial bones, for which a few present in CL6, increased to the highest number in CL4, and then decreased thereafter. In addition, lower limb and foot bones appear to be proportionally less in CL2 and CL3 compared to earlier periods.

@@

# Discussion
<!--results and chronology -->
The sampled faunal assemblage from the major residential area at Kiwulan suggests a continuation of sika deer consumption from CL6 to CL2, and then shifted to cattle in CL1. Despite deer was a substantial part of dietary practices at Kiwulan, variations in their quantity, ubiquity, butchering patterns, and proportions relative to other animals show changes that could reflect the indigenous reactions to foreign presence. Based on radiocarbon dates and diagnostic artifacts indicating specific time [@Wang2020ornament; @Chen2007], the cultural layers were further grouped into temporal sequences in relation to foreign presences for further discussion. The temporal sequences include the pre-European period (CL6, CL5), the European period (CL4), the post-European period (CL3), and the arrival of Han-Chinese immigrants (CL2, CL1). It should be noted that the European period does not necessarily mean a solid European colonial control over Kiwulan, but a broad impact from European occupation in Northern Taiwan and associated colonial economic activities on indigenous people in Yilan, such as getting involved in the long-distance trade network, tribute paying, and suffered from occasional European attacks. The small sample size after separating them into different periods might limits a comprehensive interpretation. However, comparing the results in a continous sequences still provide evident patterns reflecting changes in animal use.

@@

A consistent pattern in deer part frequencies was also identified throughout all periods. There were high frequencies of upper limb bones but low frequencies of vertebrae and ribs, suggesting selectively butchered and preferences of meaty portions. Moreover, among those meaty portions, front quarters (humerus and radius) are always more abundant than hindquarters (femur and tibia). One possible explanation is that femurs and tibiae were heavily fragmented for marrow extraction due to their high yields [@Madrigal2002; @Lupo1998], making them less identifiable in the assemblage. Marrow extraction can be supported by at least 25% of long bones showing fractures and a moderate degree of fragmentation in general. Another likelihood is that some hindquarters might be used for trading, as European records mention deer legs and meats as commodities in addition to hides [@Blusse2006, pp.220-224], but without specifying which legs. Considering the overall high proportion of meaty portions at Kiwulan, deer were likely processed not only for local consumption but also for trade.

@@

<!-- population dynamics and arguments-->
The high population during the European period inferred by pottery could contribute to the increasing animal use overall [@Wang2020pot; @Hsieh2009]. However, the impacts from foreign presence can be still identified through deer butchery patterns, relative proportions of animals, and taxon ubiquity. First, lower limbs with skinning cuts were mostly in the European period, followed by the late pre-European phase, corresponding with the rise of regional trade since the 15th century that were later intensified in the 17th century. Second, if the high frequencies of deer parts, especially heads for brains or tongues consumption, during the European period solely resulted from food stress due to population growth, a parallel rise in marrow extraction would be expected. However, long bones with fracture patterns were 10-30 % lower during the European period  and the previous phase than in other periods. Furthrmore, population growth should have led to a more proportional increase in both large and small animals. Despite a high taxonomic diversity during the European period, suids present a disproportionate rise in their ubiquity, quantity, and proportion. This indicates that deer products trade may have driven indigenous people to rely more on suids as an alternative prey, likely because of their comparable body size and meat return rates to deer [@Broughton2011prey]. Suids as a secondary resource after deer in indigenous cultures were observed in western Taiwan [@Chuang2021]. This hints the role of suids as a supplement when deer was not sufficient. After the Dutch left, deer hide trade continued but in a small scale with around 10,000 hides exported annually, which is much less than 100,000 hides in the European period [@Nakamura1997; @Tsao2011]. The small trade market and low population at Kiwulan might lead to a reduced reliance of suids, as shown by a low proportion of suids in the post-European period and eventually disappeared during the Chinese period.

<!--indigenous reaction and resilience-->
Compelling the zooarchaeological evidence and descriptions from historical records, the substantial increase in deer bones in the European period not just simply the result of population growth, but reflect intensified hunting driven by deer trade. Deer is the first priority of food source and commodities at Kiwulan since the late pre-European period and their economic benefits were intensified during the European period. It should be recognized that the role of the Dutch is both colonist who demanded tribute from Yilan, and another trader who were incorporated into the pre-existing inter-insular trade involving Han Chinese merchants and other indigenous groups in Northern Taiwan [e.g., @Torrence2000]. Local indigenous people might take a more active participation in the trade rather than resisted due to the benefits from exchanging foreign necessities (e.g. metal knives and pan) and goods (e.g. ornaments and porcelains) that were highly valued in local cultures [@Wang2020ornament]. Despite an indirect and weak Dutch colonial control in Yilan, the stress from colonial powers can not be totally neglected given the local experiences of the Dutch attacks occasionally. Everyday practices play an essential role for maintaining Indigenous traditions within a colonial setting or frequent cultural exchange [@Silliman2009; @Noe2023]. The persistence of deer consumption with similar butchery patterns over time at Kiwulan reflect a maintained tradition even after other animals were introduced such as cattle, showing community identities.

# Conclusion
By contextualizing zooarchaeological evidence into time periods relative to foreign presence, including the Europeans in the 17th century, and Chinese immigrants from the late 19th century, this study reveals the participation of indigenous groups in regional trade work. The frequencies, butchering patterns, and human modifications on deer bones confirms the active deer hides trade since the early 15th century and intensified during the European period. The results demonstrates even indigenous people in peripheral areas with relatively weak colonial control were also profoundly impacted than previously thought. In pericolonial setting, the continued prioritization of deer consumption at Kiwulan suggests a cultural persistence of food practices, while the strategic reliance on suids during the European period demonstrates adaptation to high demands for deer products, and later incorporating more muntjac in response to overhunting of sika deer. The This highlights how Kiwulan residents reacted to foreign presence and demonstrated cultural resilience. The long-term dynamics of animal use in northeastern Taiwan can enhance our understanding of Indigenous resilience in face of foreign impacts and environmental change. In addition, the role of indigenous people might be more active in the exchange network considering the large amount of foreign goods that were incorporated into local cultures and used as prestige items. A more comprehensive investigation across Indigenous settlements in northeastern Taiwan and including other dietary evidence, such as botanical remains and organic residue analysis in the future could provide a deeper insights into the regional dietary patterns and human-environment dynamics.

@@

I would like to thank the Yilan County Cultural Affairs Bureau in Taiwan for permitting access to the zooarchaeological remains used in this study and the archaeological team for the early documentation of faunal remains that provides a foundation for this study. Special thanks to Chien-Hsiang Lin at the Biodiversity Research Center, Academia Sinica, for his guidance and assistance in identifying fish bones. I also appreciate Jui-Yang Tsou for his valuable comments and assistance during the identification process. I would like to express my gratitude to Yin-Ru Chiang and my lab colleagues at the Biodiversity Research Center, Academia Sinica, for providing research space to finalize this work. I am also grateful to the reviewers of this manuscript.
